---
title: "The Effect of Nicotine Administration on Patients Determined Through Glucose Levels And Depression Tests"
output: 
  pdf_document:
    fig_caption: yes
---
 

**Authors Emily Qin\*, Sabrina Chan$\alpha$, Daniel Hu$\beta$, Mohammad Borghei$\gamma$

- $^{\ast}$ Contemporary Asian Studies Major, Health & Disease Major, Immunology Minor, University of Toronto
- $^{\alpha}$ Human Biology and CSB Major, University of Toronto
- $^{\beta}$ Bioinformatics Specialist and EEB Major, University of Toronto
- $^{\gamma}$ PharmTox specialist, University of Toronto

**Author Contributions**

- EQ, SB, DH, MB: Planned and collected data for the experiment
- EQ: Drafted Introduction
- EQ, SB: Gathered peer-reviewed research
- DH, SB, EQ: Planned Statistical Tests, edited and formatted report
- DH: Generated the R Markdown Files, Sampling and Statistical Test Code, Results and Discussion


```{r}
library(tidyverse)
```

```{r}
# reading the csv file as a df
data <- read_csv("Data Tracking - Final_Sheet.csv")
```


### Sampling Code for methods
```{r}
# Generate 80 random numbers(w/o replacement) out of the total 514 households in
# talu, and store them into a dataframe 
ran_num <- sample(1:514, 80, replace = FALSE) 
household_df <- data.frame(household_number = ran_num)

# Choosing Participants; generates a random number from 1 to user defined number
generate_random_number <- function(max_num){
  return (sample(1:max_num, 1))
}

# In case of invalid households(no one 18+, household has no members, etc),
# find Another household number that has not been chosen, and add the new number
# in place of the old number in the dataframe 
replace_num <- function(index){
   new_num <- sample(setdiff(1:max(household_df$household_number), 
                            household_df$household_number), 1)
  print(new_num)
  
  # replace the number at the vector and dataframe
  
  household_df$household_number[index] <- new_num
  
  return (household_df)
}
# Generating Treatment Groups; 1=Nicotine Treatment, 0=Placebo
# generate vector of 80 random 1's and 0's
treatment_groups <- sample(0:1, 80, replace=TRUE)

# write to csv 
#write.csv(treatment_groups, file = "treatment_groups.csv", row.names = FALSE)
```


### Displaying the Number of Participants by Smoke Group and Treatment

```{r}
# Background Info; figuring out information about the demographic
# Curate the dataframe to be more useful in future analysis

data2 <- data %>%  # Removing all participants under 40 and above 80 to 
                   # Increase power 
  filter(Age >= 40) %>%   
  filter(Age <= 80) %>% 
  filter(Smoke_Group != "HEAVY") %>%  # HEAVY Group only has one participant;
                                      # no nicotine treatment group
  rename(Blood_Glu_start = `Blood_Glu_Start(mg/dL)`) %>% 
  rename(Blood_Glu_final = `Blood_Glu_final(mg/dL)`)

# display the number of participants in each smoke category
smoke_counts <- table(data2$Smoke_Group)
nrow(data2) # get total number of participants n=45
cat("Table 1: Number of Participants in Each Smoke Category\n")
smoke_counts 

# displays the number of participants in each smoke and treatment group
participant_counts <- table(data2$Smoke_Group, data2$Treatment_Group)
cat("Table 2: Number of Participants in Each Smoke and Treatment Group\n")
print(participant_counts)

cat("\n")

# Convert Treatment_Group and Smoke_Groups to be factors for future analysis
data2$Treatment_Group_Factor <- factor(data2$Treatment_Group)
data2$Smoke_Group <- as.factor(data2$Smoke_Group)

# separate by smoking groups
non_smoker_group <- data2 %>% 
  filter(Smoke_Group == "Non")  

light_smoker_group <- data2 %>% 
  filter(Smoke_Group == "Light")  
  
moderate_smoker_group <- data2 %>% 
  filter(Smoke_Group == "Moderate")  
 
```

## Measurements 
```{r}
# separate by smoking groups
non_smoker_group <- data2 %>% 
  filter(Smoke_Group == "Non")  

light_smoker_group <- data2 %>% 
  filter(Smoke_Group == "Light")  
  
moderate_smoker_group <- data2 %>% 
  filter(Smoke_Group == "Moderate")  
 
```

# Results 

```{r}
#calculate the mean starting, mid and final glucose levels of the two treatment groups

# Calculate the mean of Blood_Glu_Start and Blood_Glu_final for each group by Treatment_Group
non_smoker_means <- aggregate(cbind(Blood_Glu_start, Blood_Glu_final) ~ 
                                Treatment_Group, data = non_smoker_group, 
                              FUN = mean)
light_smoker_means <- aggregate(cbind(Blood_Glu_start, Blood_Glu_final) ~ 
                                  Treatment_Group, data = light_smoker_group, 
                                FUN = mean)
moderate_smoker_means <- aggregate(cbind(Blood_Glu_start, Blood_Glu_final) ~ 
                                     Treatment_Group,
                                   data = moderate_smoker_group, FUN = mean)

#combine into one dataframe
means_df <- data.frame(Smoke_Group = c("Non", "Light", "Moderate"), 
                       placebo_start = c(non_smoker_means[1,2], 
                                         light_smoker_means[1,2], 
                                         moderate_smoker_means[1,2]),
                       treatment_start = c(non_smoker_means[2,2], 
                                           light_smoker_means[2,2], 
                                         moderate_smoker_means[2,2]),
                       placebo_end = c(non_smoker_means[1,3], 
                                       light_smoker_means[1,3],
                                       moderate_smoker_means[1,3] ),
                       treatment_end = c(non_smoker_means[2,3], 
                                         light_smoker_means[2,3], 
                                       moderate_smoker_means[1,3]
                                      ))
# Convert the data from wide to long format
means_df_long <- pivot_longer(means_df, 
                              cols = c(placebo_start, treatment_start, 
                                       placebo_end, treatment_end), 
                              names_to = "Treatment_Group", 
                              values_to = "Mean_Blood_Glucose")

# Reorder the smoking groups
order_smoke_group <- c("Non", "Light", "Moderate", "Heavy")

# Create a new factor variable with the desired order
means_df_long$Smoke_Group <- factor(means_df_long$Smoke_Group, 
                                     levels = order_smoke_group)

# summarizing the starting and end blood glucose levels of each group
cat("Table 3: Summary of Starting and End Blood Glucose Levels for Each Group")
print(means_df)


# Create the barplot of mean blood glucose according to smoke and treatment
ggplot(means_df_long, aes(x = Smoke_Group, y = Mean_Blood_Glucose, 
                          fill = Treatment_Group)) +
  geom_bar(stat = "identity", position = position_dodge2(preserve = "single", 
                                                         padding = 0.2)) +
  labs(x = "Smoking Group", y = "Mean Blood Glucose (mg/dL)", 
       caption = 
         "Figure 1A: Mean Blood Glucose Levels By Smoking Group and Treatment") +
  scale_fill_manual(values = c("lightblue", "Blue", "lightgreen", "green"), 
                    labels = c("Placebo - Start", "Placebo - End", 
                               "Treatment - Start", "Treatment - End")) +
  theme_bw() + 
  theme(legend.position = "right", plot.caption = element_text(hjust = 0.5, 
                                                      size = 10, face = "bold"))
```


### Two-way Anova Analysis 
```{r}
# Comparing the means between factors(treatment group and smoking levels)
# 2-way Anova analysis 

# H0: Nicotine treatment has no impact on mean blood glucose levels 
# HA: Nicotine Treatment does have an impact on mean blood glucose levels for the different smoking levels

# H0: Smoking levels have no effect on mean blood glucose levels
# HA: Smoking levels does have an effect on the mean blood glucose levels for 
# at least one treatment group 

# H0: Mean blood glucose levels does not vary depending on nicotine treatment 
# and nicotine treatment's effect on mean blood glucose levels does not vary 
# with different smoking levels 
# HA: Mean blood glucose levels is effected depending on nicotine treatment or 
# the nicotine treatment effect varies depending on smoking levels 

# store results of the two way anova 
two_way_anova <- aov(Blood_Glu_final ~ Treatment_Group * Smoke_Group, 
                     data = data2)


# actual and adjusted2-way anova results
cat("Figure 1B: Two-Way ANOVA Analysis\n")
summary(two_way_anova)

cat("Figure 1C: TukeyHSD Adjustment for Two-Way Analysis\n")
TukeyHSD(two_way_anova)
```

### 2-way Anova Analysis Assumptions
```{r}
# Checking Assumptions for 2-way anova analysis
#checking assumptions 
qqnorm(two_way_anova$residuals)
qqline(two_way_anova$residuals)

# Add caption-like text below the plot
par(mar = c(6, 4, 4, 2) + 0.1)  # Adjust bottom margin to make space for caption
mtext("Figure 1E: QQ-Plot for Blood Glucose ANOVA", side = 1, line = 5, 
      cex = 0.8, font = 2, adj = 0.5)

  
cat("Table 4: Standard Deviations of Blood Glucose Levels with Smoke and 
    Treatment Groups")
tapply(data2$Blood_Glu_final, list(data2$Treatment_Group, data2$Smoke_Group), 
       "sd")


stripchart(two_way_anova$residuals~data2$Treatment_Group*
             data2$Smoke_Group, vertical=TRUE, pch=1, 
           xlab = "Treatment and Smoking Groups(0=Placebo, 1=Nicotine)")

# add caption for strip plot 
par(mar = c(6, 4, 4, 2) + 0.1)  # Adjust bottom margin to make space for caption
mtext("Figure 1D: Strip Plot To Test Standard Variance of 
      Blood Glucose ANOVA", side = 1, line = 5, cex = 0.8, font = 2, adj = 0.5)

```


## Blood Glucose Levels and Depressive Symptoms
 
```{r}
# comparing two qualitative variables, use a scatterplot
par(mar = c(7, 4, 4, 2) + 0.1)  # Adjust bottom margin to make space for caption
plot(data2$Blood_Glu_final, data2$Depression_final, xlab = 
       "Blood Glucose Levels (mg/dl)", ylab= "Depression Score Result")

# add caption for scatterplot
mtext("Figure 2. Scatter Plot of Blood Glucose Levels and Depression Scores", 
      side = 1, line = 6, cex = 0.8, font = 2, adj = 0.5)

# adding a line of best fit to view correlation
regression <- lm(data2$Depression_final ~ data2$Blood_Glu_final)
abline(regression, col="red", lwd=2)

# extract coefficients from the regression model
intercept <- coef(regression)[1]
slope <- coef(regression)[2]

# display the equation of the line of best fit
cat("Equation of the line of best fit: y =", round(intercept, 2), "+", round(slope, 2), "x\n")

```









